---
title: "LungMAP Mouse LMEX0000004397"
subtitle: "https://www.lungmap.net/dataset/?dataset_id=LMEX0000004397&view=downloads"
author: "Patrick Ozark"
output: pdf_document
---

### Set working direcotry
```{r Working directory}
setwd("~/Projects/lungMAPmouse/")
```

# Library
```{r Library, message = FALSE, quiet = TRUE}
library(anndataR)
library(Seurat)
library(Matrix)
library(dplyr)
library(vsn)
library(DESeq2)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(ggcorrplot)
```

# Load data
```{r Load data, eval = TRUE}
# Avoid significant memory usage regenerating the data loaded here
# Loaded data is generated within this script on initial pass-through
load("./lungMAPmouse.RData")
```


# Create Seurat object using raw counts
```{r Seurat object, eval = FALSE}
raw_data <- read_h5ad("./data/LungMAP_MouseLung_CellRef.v1.1.raw.h5ad")
raw_gene_names <- raw_data$var[["features"]]
head(raw_gene_names, 10)
raw_counts <- t(as.matrix(raw_data$X))
rownames(raw_counts) <- raw_gene_names


seurat_obj <- CreateSeuratObject(
  counts = raw_counts,
  meta.data = as.data.frame(raw_data$obs),
  min.features = 250,
  min.cells = 30
)

# Verify — this should now show real gene names
head(rownames(seurat_obj), 10)
dim(raw_counts)
length(raw_gene_names)


head(seurat_obj@assays[["RNA"]]@features)
seurat_obj@assays[["RNA"]]@features["Il34", ]


# Save Seurat object
SaveSeuratRds(object = seurat_obj, file = "./data/LungMAP_Seurat-raw_obj.rds")
```

# Normalize and scale data, PCA, UMAP for Seurat object 
```{r Operations on Seurat object, eval = FALSE}
seurat.merged <- seurat_obj

## Analysis of cDNA assay only
DefaultAssay(seurat.merged) <- "RNA"

seurat.merged <- NormalizeData(seurat.merged)
seurat.merged <- FindVariableFeatures(seurat.merged)
seurat.merged <- ScaleData(seurat.merged)
seurat.merged <- RunPCA(seurat.merged, npcs = 50)
seurat.merged <- FindNeighbors(seurat.merged, dims = 1:30)
seurat.merged <- FindClusters(seurat.merged, resolution = 0.1, verbose = TRUE)
seurat.merged <- RunUMAP(seurat.merged, dims = 1:30)

options(repr.plot.width = 5, repr.plot.height = 5)
DimPlot(seurat.merged, label = TRUE, raster = TRUE) + NoLegend()

# Save Seurat.merged object to Rds
SaveSeuratRds(object = seurat.merged, file = "./data/LungMAP_Seurat-raw_merged.rds")
```

# Get AT2 cells from Seurat object
```{r Subset AT2 cells, eval = TRUE}
# Subset Seurat object (merged) for AT2 cells
seurat_AT2 <- subset(
  seurat_obj,
  subset = celltype_level3_fullname == "Alveolar type 2 cell"
)

# AT2 Pseudobulk counts by timepoint (orig.ident)
at2_pseudo_counts <- AggregateExpression(
  object = seurat_AT2,
  group.by = "orig.ident",
  assays = "RNA",
  slot = "counts"
)
at2_matrix_timepoint <- as.matrix(at2_pseudo_counts$RNA)

# AT2 Pseudobulk counts by sample (DataID)
at2_pseudo_counts_s <- AggregateExpression(
  object = seurat_AT2,
  group.by = "DataID",
  assays = "RNA",
  slot = "counts"
)
at2_m_s <- as.matrix(at2_pseudo_counts_s$RNA)

# Read in metadata .txt file
cell_meta <- read.table("data/Cell.Metadata.txt", header = TRUE, sep = "\t") %>% data.frame(stringsAsFactors = FALSE)
rownames(cell_meta) <- cell_meta$Cell.ID

```

# DESeq2
```{r DESeq2 on AT2 cells, eval = FALSE}
# Read in metadata needed for DESeq2
deseq_metadata_timepoint <- read.csv("./data/LungMAP-deseq2-metadata_timepoint.csv")

at2_counts <- at2_matrix_timepoint
head(at2_counts)

# Construct DESeq2 dataset object
at2_dds <- DESeqDataSetFromMatrix(
  countData = at2_counts,
  colData = deseq_metadata_timepoint,
  design = ~age,
  tidy = FALSE # Would be TRUE if rownames were not already gene names and were instead present as the first column in the counts matrix
)

at2_dds <- DESeq(at2_dds)
# Save RDS of AT2 DESeq data set
saveRDS(at2_dds, "./data/AT2_DESeq2-dds.rds")

at2_results <- results(at2_dds, tidy = TRUE)
head(at2_results)

summary(at2_results)
```

# rlog-normalize AT2 cell expression counts
```{r rlog-normalization of DEseq2 counts, eval = FALSE}
at2_dds <- readRDS("./data/AT2_DESeq2-dds.rds")
# rlog transform the DESeq2 dataset: Remove the dependence of the variance on the mean, particularly the high variance of the logarithm of count data when the mean is low. There will be some elevated standard deviation in the lower count range
rld_at2 <- rlog(at2_dds, blind = FALSE) # blind=FALSE since design specifies 'age' as comparison group
head(assay(rld_at2)) # assay pulls out normalized values
```

# SD and PCA
```{r AT2 cells PCA, eval = TRUE}
# Mean standard deviation plot
meanSdPlot(assay(rld_at2))

# PCA plot
rlog_at2_PCAData <- plotPCA(rld_at2, intgroup = "age", returnData = TRUE)

# Include percent of the variance explained by the top 2 principal components
percentVar <- round(100 * attr(rlog_at2_PCAData, "percentVar"), 2)

ggplot(rlog_at2_PCAData, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 2) +
  geom_text(aes(label = name), vjust = 1.5, size = 3, show.legend = FALSE) +
  labs(
    title = "LungMAP PCA",
    x = paste0("PC1: ", percentVar[1], "%"),
    y = paste0("PC2: ", percentVar[2], "%")
  ) +
  theme_bw()
```

# AT2 cells Correlation (Spearman) calculation by timepoint (8 'samples')
```{r AT2 cells correlation for all genes, eval = TRUE}
# Overall correlation global gene expression
rld_at2_matrix <- assay(rld_at2)
rld_at2_cor <- cor(rld_at2_matrix, method = "spearman")
pheatmap(rld_at2_cor)
```

```{r AT2 cell gene expression correlation with IL34, eval = FALSE}
# rlog expression matrix for AT2 cells
expr <- rld_at2_matrix

target_gene <- "Il34"

# Extract target vector (as numeric)
target_vec <- as.numeric(expr[target_gene, ])
n_samples <- length(target_vec)

# Prepare output storage
genes <- rownames(expr)
n_genes <- length(genes)
results <- data.frame(gene = genes, rho = NA_real_, p.value = NA_real_, stringsAsFactors = FALSE)


# Loop through genes and compute Spearman correlation with Il34
for (i in seq_len(n_genes)) {
  g <- genes[i]
  # skip if same gene
  if (g == target_gene) {
    results$rho[i] <- 1
    results$p.value[i] <- 0
    next
  }
  # get numeric vector for gene g
  vec <- as.numeric(expr[g, ])
  # Check for enough non-NA pairs
  ok <- complete.cases(vec, target_vec)
  if (sum(ok) < 3) { # too few data points to compute correlation reliably
    results$rho[i] <- NA
    results$p.value[i] <- NA
    next
  }
  # compute Spearman correlation using cor.test
  ct <- suppressWarnings(cor.test(vec[ok], target_vec[ok], method = "spearman", exact = FALSE))
  results$rho[i] <- unname(ct$estimate) # Spearman's rho
  results$p.value[i] <- ct$p.value
}

# Post-process: remove rows with NA rho (if you want(optional), and sort by rho descending
# results <- results[!is.na(results$rho), ]
results <- results[order(-results$rho), ]

# Save to CSV
write.csv(results, "./data/AT2_Il34_spearman_by_gene.csv", row.names = FALSE)
```

## AT2 cells IL34 scatterplots vs every other gene (Spearman + best fit line)
```{r AT2 cell scatter plots, eval = FALSE}
expr <- rld_at2_matrix
# Sample names
sample_names <- colnames(expr)
if (is.null(sample_names) || any(sample_names == "")) {
  sample_names <- paste0("Sample", seq_len(ncol(expr)))
}

# Output folder
plot_dir <- "./plots"
pdf_file <- file.path(plot_dir, "AT2_Il34_vs_all_genes_spearman_scatter.pdf")

# Genes to plot: all except Il34, and drop genes with NA rho from your results table
genes_to_plot <- results$gene[results$gene != target_gene]
genes_to_plot <- genes_to_plot[!is.na(results$rho[match(genes_to_plot, results$gene)])]

# Limit number of plots
max_plots <- 1000
genes_to_plot <- head(genes_to_plot, max_plots)

# Create a single multi-page PDF
pdf(pdf_file, width = 6, height = 5, onefile = TRUE)

for (g in genes_to_plot) {
  x <- as.numeric(expr[g, ])
  y <- target_vec

  df <- data.frame(
    sample = sample_names,
    other_gene = x,
    Il34 = y,
    stringsAsFactors = FALSE
  )
  df <- df[complete.cases(df$other_gene, df$Il34), , drop = FALSE]

  # Spearman correlation (for annotation)
  ct <- suppressWarnings(cor.test(df$other_gene, df$Il34, method = "spearman", exact = FALSE))
  rho <- unname(ct$estimate)
  pval <- ct$p.value

  # Linear model for the plotted best-fit line + r^2
  fit <- tryCatch(lm(Il34 ~ other_gene, data = df), error = function(e) NULL)
  r2 <- if (!is.null(fit)) summary(fit)$r.squared else NA_real_

  # Format stats
  rho_txt <- sprintf("%.3f", rho)
  p_txt <- format.pval(pval, digits = 3, eps = 1e-300)
  r2_txt <- if (is.na(r2)) "NA" else sprintf("%.3f", r2)

  # Scatter + best-fit line + Spearman + r^2
  p <- ggplot(df, aes(x = other_gene, y = Il34)) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", se = FALSE) +
    # Label points (8 samples):
    geom_text(aes(label = sample), vjust = -0.7, size = 3) +
    labs(
      title = paste0("Il34 vs ", g),
      x = paste0(g, " (rlog expression)"),
      y = "Il34 (rlog expression)",
      subtitle = paste0(
        "Spearman rho = ", rho_txt,
        ", p = ", p_txt,
        " | r^2 ", r2_txt
      )
    ) +
    theme_bw()
  print(p)
}

dev.off()
```

```{r GSEA on top correlated genes with IL34, eval = TRUE}
library(clusterProfiler)
library(org.Mm.eg.db)
# If you have a ranked list of genes (e.g., from differential expression),
# create a ranked gene list from correlation results
# using the Spearman correlation with Il34 from earlier analysis

# Read the correlation results
il34_cor_results <- read.csv("./data/AT2_Il34_spearman_by_gene.csv",
  stringsAsFactors = FALSE
)

# Create ranked list for GSEA
ranked_genes <- il34_cor_results %>%
  dplyr::select(gene, rho) %>%
  na.omit() %>%
  arrange(desc(rho))

# Convert to named vector for GSEA
ranked_vector <- setNames(ranked_genes$rho, ranked_genes$gene)

# Run GSEA with GO terms
gsea_result <- gseGO(
  geneList = ranked_vector,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "SYMBOL",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

# Plot GSEA results
pdf("./plots/AT2_Il34_GSEA_dotplot.pdf", width = 10, height = 8)
dotplot(gsea_result, showCategory = 15) +
  ggtitle("GSEA: Gene Sets Correlated with Il34 Expression")
dev.off()

# Ridge plot for enriched pathways
pdf("./plots/AT2_Il34_GSEA_ridgeplot.pdf", width = 10, height = 8)
ridgeplot(gsea_result, showCategory = 15) +
  ggtitle("GSEA Ridge Plot: Il34-Correlated Pathways")
dev.off()
```

# Cluster AT2 cell rlog(expression) matrix using timepoints
```{r Cluster AT2 cells, eval = FALSE}
## Hierarchical clustering of genes (rows) in a genes x samples matrix
## - Produces: (1) CSV with gene -> cluster assignment, (2) PDF heatmap with dendrogram(s)

# Options
expr <- rld_at2_matrix # <--- set this to your matrix object (genes=rows, samples=cols)
k <- 15 # number of gene clusters to cut the gene dendrogram into
linkage_method <- "ward.D2" # "complete", "average", "single", "ward.D2", ...
distance_metric <- "euclidean" # "euclidean" or "correlation"
cor_method <- "spearman" # used only if distance_metric == "correlation": "pearson" or "spearman"

scale_rows <- TRUE # TRUE = z-score each gene across samples before clustering/plotting, so clustering REFLECTS EXPRESSION PATTERNS ACROSS SAMPLES, NOT EXPRESSION MAGNITUDE
cluster_samples <- FALSE # TRUE = also cluster samples (columns) and show sample dendrogram


# Prep data
mat <- as.matrix(expr)
stopifnot(!is.null(rownames(mat)), !is.null(colnames(mat)))

# Optional: row-scale (z-score) genes across samples
mat_use <- if (scale_rows) t(scale(t(mat))) else mat

# Drop genes with zero/undefined variance (helps correlation + scaling behave)
gene_sd <- apply(mat_use, 1, sd, na.rm = TRUE)
keep <- is.finite(gene_sd) & gene_sd > 0
mat_use <- mat_use[keep, , drop = FALSE]

if (k > nrow(mat_use)) stop("k is larger than the number of genes after filtering.")

# Cluster genes
if (distance_metric == "euclidean") {
  d_genes <- dist(mat_use, method = "euclidean")
} else if (distance_metric == "correlation") {
  cor_genes <- cor(t(mat_use), method = cor_method, use = "pairwise.complete.obs")
  d_genes <- as.dist(1 - cor_genes)
} else {
  stop("distance_metric must be 'euclidean' or 'correlation'.")
}

hc_genes <- hclust(d_genes, method = linkage_method)
gene_clusters <- cutree(hc_genes, k = k)

# Save gene cluster assignments
cluster_df <- data.frame(gene = rownames(mat_use), cluster = gene_clusters)
write.csv(cluster_df, paste0("./data/AT2_gene_hclust", "_clusters_k", k, "_dist-", distance_metric, ".csv"), row.names = FALSE)

# Sample clustering (If specified in options)
hc_samples <- FALSE
if (cluster_samples) {
  if (distance_metric == "euclidean") {
    d_samples <- dist(t(mat_use), method = "euclidean")
  } else {
    cor_samples <- cor(mat_use, method = cor_method, use = "pairwise.complete.obs")
    d_samples <- as.dist(1 - cor_samples)
  }
  hc_samples <- hclust(d_samples, method = linkage_method)
}

# Heatmap + dendrogram
annotation_row <- data.frame(Cluster = factor(gene_clusters))
rownames(annotation_row) <- rownames(mat_use)

# Order clusters by the order in which they appear in the dendrogram to aid in visual interpretation
row_order <- hc_genes$order
ordered_genes <- rownames(mat_use)[row_order]
clusters_in_order <- gene_clusters[ordered_genes]

# Keep only first occurrence of each cluster
cluster_levels_ordered <- unique(clusters_in_order)

# Reorder the cluster
annotation_row$Cluster <- factor(
  gene_clusters,
  levels = cluster_levels_ordered
)

# Make gene cluster named by gene IDs
names(gene_clusters) <- rownames(mat_use)

# Set scale colors
clrsp <- colorRampPalette(c("blue", "white", "red"))
clrs <- clrsp(200)

# Custom scale
breaks <- seq(-2, 2, length.out = 201)

outfile <- paste0("./plots/AT2_gene_hclust", "_heatmap_k", k, "_dist-", distance_metric, ".pdf")
pdf(outfile, width = 7, height = 9)
pheatmap(
  mat_use,
  cluster_rows = hc_genes,
  cluster_cols = hc_samples, # FALSE if cluster_samples == FALSE
  cutree_rows = k, # visually separates the k gene clusters
  annotation_row = annotation_row,
  color = clrs,
  breaks = breaks,
  show_rownames = FALSE, # set TRUE if you have few genes and want names shown
  main = paste0(
    "Gene hclust (k=", k,
    ", linkage=", linkage_method,
    ", dist=", distance_metric,
    if (distance_metric == "correlation") paste0(" [", cor_method, "]") else "",
    if (scale_rows) ", row-zscore" else ""
  )
)
dev.off()
```

## Cluster analysis
```{r Isolate AT2 cluster 6, eval = TRUE}
clusters <- read.table("./data/AT2_gene_hclust_clusters_k12.csv", sep = ",", header = TRUE)
cluster_6 <- subset(clusters, cluster == 6)
cluster6_genes <- cluster_6[, 1]
# write.table(cluster6_genes, "./data/AT2_cluster6_genes.txt", row.names = FALSE, quote = FALSE, col.names = FALSE)

# Subset rld_at2_matrix to cluster 6 genes only
cluster6_genes <- as.character(cluster_6$gene)
cluster6_matrix <- rld_at2_matrix[cluster6_genes, , drop = FALSE]
cluster6_gsea_m <- as.data.frame(cluster6_matrix)

cluster6_gsea_m["description"] <- as.character("na")
cluster6_gsea_m["name"] <- row.names(cluster6_gsea_m)
new_order_pos <- c(10, 9, 1:8)
cluster6_gsea_m <- cluster6_gsea_m[, new_order_pos]

# write.table(cluster6_gsea_m, "./data/gsea_at2_cluster6_rlogGeneCounts.txt", sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
```

```{r Cluster heatmap, eval = TRUE}
library(ComplexHeatmap)
library(circlize)

# Create heatmap for cluster 6 genes
# Using z-score scaling for better visualization
cluster6_scaled <- t(scale(t(cluster6_matrix)))

# Define color palette
heat_colors <- colorRamp2(
  c(-2, 0, 2),
  c("#2166AC", "white", "#B2182B")
)

ComplexHeatmap::Heatmap(
  cluster6_scaled,
  name = "Z-score",
  col = heat_colors,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_title = "Cluster 6 Genes Across Timepoints",
  row_title = paste0("Genes (n=", nrow(cluster6_scaled), ")"),
  clustering_distance_rows = "spearman",
  clustering_method_rows = "ward.D2",
  cluster_columns = FALSE
)
```

```{r Cluster GO Enrichment analysis, eval = TRUE}
# Convert gene symbols to Entrez IDs for cluster 6
cluster6_entrez <- bitr(
  cluster6_genes,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

# Remove NAs
cluster6_entrez <- na.omit(cluster6_entrez)

# GO Biological Process enrichment
go_bp <- enrichGO(
  gene = cluster6_entrez$ENTREZID,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE
)

# GO Molecular Function enrichment
go_mf <- enrichGO(
  gene = cluster6_entrez$ENTREZID,
  OrgDb = org.Mm.eg.db,
  ont = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE
)

# GO Cellular Component enrichment
go_cc <- enrichGO(
  gene = cluster6_entrez$ENTREZID,
  OrgDb = org.Mm.eg.db,
  ont = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE
)

# Save GO results
write.csv(as.data.frame(go_bp), "./data/AT2_cluster6_GO_BP_results.csv", row.names = FALSE)
write.csv(as.data.frame(go_mf), "./data/AT2_cluster6_GO_MF_results.csv", row.names = FALSE)
write.csv(as.data.frame(go_cc), "./data/AT2_cluster6_GO_CC_results.csv", row.names = FALSE)

# Plot top GO terms
pdf("./plots/AT2_cluster6_GO_BP_dotplot.pdf", width = 10, height = 8)
dotplot(go_bp, showCategory = 15) + # Use orderBy to control ordering by gene ratio or p value
  ggtitle("Cluster 6: Top GO Biological Process Terms")
dev.off()

pdf("./plots/AT2_cluster6_GO_BP_barplot.pdf", width = 10, height = 8)
barplot(go_bp, showCategory = 15) +
  ggtitle("Cluster 6: Top GO Biological Process Terms")
dev.off()

pdf("./plots/AT2_cluster6_GO_MF_dotplot.pdf", width = 10, height = 8)
dotplot(go_mf, showCategory = 15) +
  ggtitle("Cluster 6: Top GO Molecular Function Terms")
dev.off()

pdf("./plots/AT2_cluster6_GO_MF_barplot.pdf", width = 10, height = 8)
barplot(go_mf, showCategory = 15) +
  ggtitle("Cluster 6: Top GO Molecular Function Terms")
dev.off()

pdf("./plots/AT2_cluster6_GO_CC_dotplot.pdf", width = 10, height = 8)
dotplot(go_cc, showCategory = 15) +
  ggtitle("Cluster 6: Top GO Cellular Component Terms")
dev.off()

pdf("./plots/AT2_cluster6_GO_CC_barplot.pdf", width = 10, height = 8)
barplot(go_cc, showCategory = 15) +
  ggtitle("Cluster 6: Top GO Cellular Component Terms")
dev.off()
```

```{r Cluster KEGG pathway enrichment, eval = TRUE}
# KEGG pathway enrichment
kegg_enrich <- enrichKEGG(
  gene = cluster6_entrez$ENTREZID,
  organism = "mmu", # Mouse
  keyType = "kegg",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)

# Convert KEGG IDs to gene symbols for readability
kegg_enrich <- setReadable(kegg_enrich, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")

# Save KEGG results
write.csv(as.data.frame(kegg_enrich), "./data/AT2_cluster6_KEGG_results.csv", row.names = FALSE)


# Plot KEGG pathways
if (nrow(as.data.frame(kegg_enrich)) > 0) {
  pdf("./plots/AT2_cluster6_KEGG_dotplot.pdf", width = 10, height = 8)
  dotplot(kegg_enrich, showCategory = 15) +
    ggtitle("Cluster 6: KEGG Pathway Enrichment")
  dev.off()
}

# Visualize specific pathway (example: if enriched)
# Replace "mmu04060" with actual enriched pathway ID
# pathview(gene.data = cluster6_entrez$ENTREZID,
#          pathway.id = "mmu04060",
#          species = "mmu",
#          out.suffix = "cluster6")
```

```{r Cluster Reactome pathway enrichment, eval = FALSE}
# Reactome pathway enrichment (alternative pathway database)
# library(ReactomePA)
# reactome_enrich <- enrichPathway(
#   gene = cluster6_entrez$ENTREZID,
#   organism = "mouse",
#   pvalueCutoff = 0.05,
#   pAdjustMethod = "BH",
#   qvalueCutoff = 0.2,
#   readable = TRUE
# )
```


```{r Cluster gene correlation network analysis, eval = TRUE}
# Build gene-gene correlation network for cluster 6 genes
cor_matrix <- cor(t(cluster6_matrix), method = "spearman")

# Set diagonal to 0
diag(cor_matrix) <- 0

# Filter for strong correlations (|r| > 0.7)
threshold <- 0.7
cor_matrix_filtered <- cor_matrix
cor_matrix_filtered[abs(cor_matrix_filtered) < threshold] <- 0

# Create adjacency matrix for network analysis
adj_matrix <- abs(cor_matrix_filtered)

# Plot correlation heatmap
pdf("./plots/AT2_cluster6_gene_correlation_heatmap.pdf", width = 12, height = 12)
pheatmap(
  cor_matrix[seq_len(min(50, nrow(cor_matrix))), seq_len(min(50, ncol(cor_matrix)))],
  color = colorRampPalette(c("blue", "white", "red"))(100),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  main = "Cluster 6 Gene-Gene Correlation Matrix (Top 50 Genes)"
)
dev.off()
```

```{r Cluster PCA, eval = TRUE}
# PCA on cluster 6 genes
pca_result <- prcomp(t(cluster6_matrix), center = TRUE, scale. = TRUE)

# Get variance explained
var_explained <- summary(pca_result)$importance[2, ] * 100

# Create PCA data frame
pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Sample = rownames(pca_result$x),
  Age = colData(rld_at2)$age[match(rownames(pca_result$x), colnames(rld_at2_matrix))]
)

# Plot PCA
pdf("./plots/AT2_cluster6_PCA.pdf", width = 8, height = 6)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Age)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -1, size = 3) +
  labs(
    title = "PCA: Cluster 6 Gene Expression",
    x = paste0("PC1 (", round(var_explained[1], 1), "% variance)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "% variance)")
  ) +
  scale_color_manual(values = c("P7" = "#4DAF4A", "P14" = "#377EB8", "P21" = "#E41A1C")) +
  theme_bw()
dev.off()

# Loadings plot (top contributing genes)
loadings_df <- data.frame(
  Gene = rownames(pca_result$rotation),
  PC1_loading = abs(pca_result$rotation[, 1]),
  PC2_loading = abs(pca_result$rotation[, 2])
) %>%
  arrange(desc(PC1_loading))

# No need to execute this for now
# write.csv(loadings_df[1:50, ], "./data/AT2_cluster6_PCA_top_loadings.csv", row.names = FALSE)
```


