---
title: "LungMAP Mouse LMEX0000004397 Sample-level Analysis"
subtitle: "https://www.lungmap.net/dataset/?dataset_id=LMEX0000004397&view=downloads"
author: "Patrick Ozark"
output: pdf_document
---

### Set working direcotry
```{r Working directory}
setwd("~/Projects/lungMAPmouse/")
```

# Library
```{r Library, message = FALSE, quiet = TRUE}
library(Seurat)
library(Matrix)
library(dplyr)
library(vsn)
library(DESeq2)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(ggcorrplot)
```

# Load data
```{r Load data, eval = TRUE}
# Avoid significant memory usage regenerating the data loaded here
# Loaded data is generated within this script on initial pass-through
load("./lungMAPmouse.RData")
```

# Get AT2 cells from Seurat object
```{r Subset AT2 cells, eval = TRUE}
# Subset Seurat object (merged) for AT2 cells
seurat_AT2 <- subset(
  seurat_obj,
  subset = celltype_level3_fullname == "Alveolar type 2 cell"
)

# AT2 Pseudobulk counts by sample (DataID)
at2_pseudo_counts_s <- AggregateExpression(
  object = seurat_AT2,
  group.by = c("orig.ident","DataID"),
  assays = "RNA",
  slot = "counts"
)
at2_m_s <- as.matrix(at2_pseudo_counts_s$RNA)

# Read in metadata .txt file
cell_meta <- read.table("data/Cell.Metadata.txt", header = TRUE, sep = "\t") %>% data.frame(stringsAsFactors = FALSE)
rownames(cell_meta) <- cell_meta$Cell.ID

```

# DESeq2
```{r DESeq2 on AT2 cells, eval = TRUE}
# Read in metadata needed for DESeq2
deseq_metadata_sample <- read.csv("./data/LungMAP-deseq2-metadata_sample.csv")

at2_counts <- at2_m_s
head(at2_counts)

# Construct DESeq2 dataset object
at2_dds <- DESeqDataSetFromMatrix(
  countData = at2_counts,
  colData = deseq_metadata_sample,
  design = ~age,
  tidy = FALSE # Would be TRUE if rownames were not already gene names and were instead present as the first column in the counts matrix
)

at2_dds <- DESeq(at2_dds)
# Save RDS of AT2 DESeq data set
saveRDS(at2_dds, "./data/AT2_DESeq2-dds_sample.rds")

at2_results <- results(at2_dds, tidy = TRUE)
head(at2_results)

summary(at2_results)
```

# rlog-normalize AT2 cell expression counts
```{r rlog-normalization of DEseq2 counts, eval = TRUE}
at2_dds_s <- readRDS("./data/AT2_DESeq2-dds_sample.rds")
# rlog transform the DESeq2 dataset: Remove the dependence of the variance on the mean, particularly the high variance of the logarithm of count data when the mean is low. There will be some elevated standard deviation in the lower count range
rld_at2_s <- rlog(at2_dds_s, blind = FALSE) # blind=FALSE since design specifies 'age' as comparison group
head(assay(rld_at2_s)) # assay pulls out normalized values
```

# SD and PCA
```{r AT2 cells PCA, eval = TRUE}
# Mean standard deviation plot
meanSdPlot(assay(rld_at2_s))

# PCA plot
rlog_at2_s_PCAData <- plotPCA(rld_at2_s, intgroup = "age", returnData = TRUE)

# Include percent of the variance explained by the top 2 principal components
percentVar <- round(100 * attr(rlog_at2_s_PCAData, "percentVar"), 2)

pdf("./plots/AT2_samples_PCA.pdf", width = 8, height = 6)
ggplot(rlog_at2_s_PCAData, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 2) +
  geom_text(aes(label = name), vjust = 1.5, size = 3, show.legend = FALSE) +
  labs(
    title = "LungMAP Samples PCA",
    x = paste0("PC1: ", percentVar[1], "%"),
    y = paste0("PC2: ", percentVar[2], "%")
  ) +
  theme_bw()
dev.off()
```

# AT2 cells Correlation (Spearman) calculation by sample (16 samples)
```{r AT2 cells correlation for all genes, eval = TRUE}
# Overall correlation global gene expression
rld_at2_m_s <- assay(rld_at2_s)
rld_at2_cor_s <- cor(rld_at2_m_s, method = "spearman")
pheatmap(rld_at2_cor_s)
```

```{r AT2 cell gene expression correlation with IL34, eval = TRUE}
# rlog expression matrix for AT2 cells and samples
expr <- rld_at2_m_s

target_gene <- "Il34"

# Extract target vector (as numeric)
target_vec <- as.numeric(expr[target_gene, ])
n_samples <- length(target_vec)

# Prepare output storage
genes <- rownames(expr)
n_genes <- length(genes)
results <- data.frame(gene = genes, rho = NA_real_, p.value = NA_real_, stringsAsFactors = FALSE)


# Loop through genes and compute Spearman correlation with Il34
for (i in seq_len(n_genes)) {
  g <- genes[i]
  # skip if same gene
  if (g == target_gene) {
    results$rho[i] <- 1
    results$p.value[i] <- 0
    next
  }
  # get numeric vector for gene g
  vec <- as.numeric(expr[g, ])
  # Check for enough non-NA pairs
  ok <- complete.cases(vec, target_vec)
  if (sum(ok) < 3) { # too few data points to compute correlation reliably
    results$rho[i] <- NA
    results$p.value[i] <- NA
    next
  }
  # compute Spearman correlation using cor.test
  ct <- suppressWarnings(cor.test(vec[ok], target_vec[ok], method = "spearman", exact = FALSE))
  results$rho[i] <- unname(ct$estimate) # Spearman's rho
  results$p.value[i] <- ct$p.value
}

# Post-process: remove rows with NA rho (if you want(optional), and sort by rho descending
# results <- results[!is.na(results$rho), ]
results <- results[order(-results$rho), ]

# Save to CSV
write.csv(results, "./data/AT2_samples_Il34_spearman_by_gene.csv", row.names = FALSE)
```

## AT2 cells IL34 scatterplots vs every other gene (Spearman + best fit line) by sample
```{r AT2 cell scatter plots, eval = TRUE}
expr <- rld_at2_m_s
# Sample names
sample_names <- colnames(expr)
if (is.null(sample_names) || any(sample_names == "")) {
  sample_names <- paste0("Sample", seq_len(ncol(expr)))
}

# Output folder
plot_dir <- "./plots"
pdf_file <- file.path(plot_dir, "AT2_samples_Il34_vs_all_genes_spearman_scatter.pdf")

# Genes to plot: all except Il34, and drop genes with NA rho from your results table
genes_to_plot <- results$gene[results$gene != target_gene]
genes_to_plot <- genes_to_plot[!is.na(results$rho[match(genes_to_plot, results$gene)])]

# Limit number of plots
max_plots <- 1000
genes_to_plot <- head(genes_to_plot, max_plots)

# Create a single multi-page PDF
pdf(pdf_file, width = 6, height = 5, onefile = TRUE)

for (g in genes_to_plot) {
  x <- as.numeric(expr[g, ])
  y <- target_vec

  df <- data.frame(
    sample = sample_names,
    other_gene = x,
    Il34 = y,
    stringsAsFactors = FALSE
  )
  df <- df[complete.cases(df$other_gene, df$Il34), , drop = FALSE]

  # Spearman correlation (for annotation)
  ct <- suppressWarnings(cor.test(df$other_gene, df$Il34, method = "spearman", exact = FALSE))
  rho <- unname(ct$estimate)
  pval <- ct$p.value

  # Linear model for the plotted best-fit line + r^2
  fit <- tryCatch(lm(Il34 ~ other_gene, data = df), error = function(e) NULL)
  r2 <- if (!is.null(fit)) summary(fit)$r.squared else NA_real_

  # Format stats
  rho_txt <- sprintf("%.3f", rho)
  p_txt <- format.pval(pval, digits = 3, eps = 1e-300)
  r2_txt <- if (is.na(r2)) "NA" else sprintf("%.3f", r2)

  # Scatter + best-fit line + Spearman + r^2
  p <- ggplot(df, aes(x = other_gene, y = Il34)) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", se = FALSE) +
    # Label points (8 samples):
    geom_text(aes(label = sample), vjust = -0.7, size = 3) +
    labs(
      title = paste0("Il34 vs ", g),
      x = paste0(g, " (rlog expression)"),
      y = "Il34 (rlog expression)",
      subtitle = paste0(
        "Spearman rho = ", rho_txt,
        ", p = ", p_txt,
        " | r^2 ", r2_txt
      )
    ) +
    theme_bw()
  print(p)
}

dev.off()
```

## GSEA on top correlated genes with IL34 expression in AT2 cells
```{r GSEA on top correlated genes with IL34, eval = TRUE}
library(clusterProfiler)
library(org.Mm.eg.db)
# If you have a ranked list of genes (e.g., from differential expression),
# create a ranked gene list from correlation results
# using the Spearman correlation with Il34 from earlier analysis

# Read the correlation results
il34_cor_results <- read.csv("./data/AT2_samples_Il34_spearman_by_gene.csv",
  stringsAsFactors = FALSE
)

# Create ranked list for GSEA
ranked_genes <- il34_cor_results %>%
  dplyr::select(gene, rho) %>%
  na.omit() %>%
  arrange(desc(rho))

# Convert to named vector for GSEA
ranked_vector <- setNames(ranked_genes$rho, ranked_genes$gene)

# Run GSEA with GO terms for genes correlated with IL34 expression
gsea_result <- gseGO(
  geneList = ranked_vector,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  keyType = "SYMBOL",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  eps = 0 # Don't estimate p-value for higher accuracy
)

# Plot GSEA results
pdf("./plots/AT2_samples_Il34_GSEA_dotplot.pdf", width = 10, height = 8)
dotplot(gsea_result, showCategory = 15) + # Use orderBy to control ordering by gene ratio or p value
  ggtitle("GSEA: Gene Sets Correlated with Il34 Expression")
dev.off()

# Ridge plot for enriched pathways
pdf("./plots/AT2_samples_Il34_GSEA_ridgeplot.pdf", width = 10, height = 8)
ridgeplot(gsea_result, showCategory = 15) +
  ggtitle("GSEA Ridge Plot: Il34-Correlated Pathways")
dev.off()
```

# Plots of IL34 expression for each sample across timepoints
```{r }
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(tibble)

corr <- read_csv("./data/AT2_samples_Il34_spearman_by_gene.csv")

# Remove rows with NA correlations
corr <- corr %>%
  filter(!if_any(everything(), is.na))

# Extract gene list
genes_to_plot <- corr$gene

# Keep only genes present in expression matrix
genes_to_plot <- intersect(genes_to_plot, rownames(rld_at2_m_s))


# Convert to long format for plotting
expr_long <- rld_at2_m_s[genes_to_plot, ] %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(
    -gene,
    names_to = "DataID",
    values_to = "expression"
  )


# Get meta data and match the DataID (sample names) with their timepoint (orig.ident)
meta <- seurat_obj@meta.data %>%
  dplyr::select(DataID, orig.ident) %>%
  distinct()

# Format sample names obtained from seurat object to enable joining of the meta and long-format expression data
meta$DataID <- str_replace_all(meta$DataID, "_", "-")

# Left join to retain all data points from long-format expression data
expr_long <- expr_long %>%
  left_join(meta, by = "DataID")


# Order timepoints
expr_long$orig.ident <- factor(
  expr_long$orig.ident,
  levels = sort(unique(expr_long$orig.ident))
)


# Calculate mean and sd
summary_df <- expr_long %>%
  group_by(gene, orig.ident) %>%
  summarize(
    mean = mean(expression),
    sd = sd(expression),
    .groups = "drop"
  )

# Ensure timepoints are ordered
expr_long$orig.ident <- factor(expr_long$orig.ident,
                        levels = unique(expr_long$orig.ident))

# Set number of plots
max_plots <- Inf
genes_to_plot <- head(ranked_genes, max_plots)$gene # Careful with genes_to_plot, it is a variable whose value is changed more than once depending on analysis context

pdf("./plots/AT2_sample-expr_across-timepoints_all-genes.pdf", width = 6, height = 5)
# for (g in unique(expr_long$gene)) {
for (g in genes_to_plot) {
  plot_df <- expr_long %>% filter(gene == g)
  plot_summary_df <- summary_df %>% filter(gene == g)

  p <- ggplot(plot_df, aes(x = orig.ident, y = expression)) +
    geom_point(size = 1,color = "blue") +
    labs(
      title = g,
      x = "Timepoint",
      y = "rlog(pseudobulk expression)"
    ) +
    theme_bw() +
    
    geom_errorbar(data = plot_summary_df, aes(x = orig.ident, 
                                              ymin = mean - sd, 
                                              ymax = mean + sd),
                  inherit.aes = FALSE,
                  width = 0.15,
                  linewidth = 0.20) +
    geom_line(data = plot_summary_df, aes(y = mean,
                                   group = gene),
              linewidth = 0.20)
  

print(p)
}

dev.off()

```

# Cluster AT2 cell rlog(expression) matrix
```{r Cluster AT2 cells, eval = FALSE}
## Hierarchical clustering of genes (rows) in a genes x samples matrix
## - Produces: (1) CSV with gene -> cluster assignment, (2) PDF heatmap with dendrogram(s)

# Options
expr <- rld_at2_m_s # <--- set this to your matrix object (genes=rows, samples=cols)
k <- 12 # number of gene clusters to cut the gene dendrogram into
linkage_method <- "ward.D2" # "complete", "average", "single", "ward.D2", ...
distance_metric <- "euclidean" # "euclidean" or "correlation"
cor_method <- "spearman" # used only if distance_metric == "correlation": "pearson" or "spearman"

scale_rows <- TRUE # TRUE = z-score each gene across samples before clustering/plotting, so clustering REFLECTS EXPRESSION PATTERNS ACROSS SAMPLES, NOT EXPRESSION MAGNITUDE
cluster_samples <- FALSE # TRUE = also cluster samples (columns) and show sample dendrogram


# Prep data
mat <- as.matrix(expr)
stopifnot(!is.null(rownames(mat)), !is.null(colnames(mat)))

# Optional: row-scale (z-score) genes across samples
mat_use <- if (scale_rows) t(scale(t(mat))) else mat

# Drop genes with zero/undefined variance (helps correlation + scaling behave)
gene_sd <- apply(mat_use, 1, sd, na.rm = TRUE)
keep <- is.finite(gene_sd) & gene_sd > 0
mat_use <- mat_use[keep, , drop = FALSE]

if (k > nrow(mat_use)) stop("k is larger than the number of genes after filtering.")

# Cluster genes
if (distance_metric == "euclidean") {
  d_genes <- dist(mat_use, method = "euclidean")
} else if (distance_metric == "correlation") {
  cor_genes <- cor(t(mat_use), method = cor_method, use = "pairwise.complete.obs")
  d_genes <- as.dist(1 - cor_genes)
} else {
  stop("distance_metric must be 'euclidean' or 'correlation'.")
}

hc_genes <- hclust(d_genes, method = linkage_method)
gene_clusters <- cutree(hc_genes, k = k)

# Save gene cluster assignments
cluster_df <- data.frame(gene = rownames(mat_use), cluster = gene_clusters)
write.csv(cluster_df, paste0("./data/AT2-samples_gene_hclust", "_clusters_k", k,"_dist-", distance_metric, ".csv"), row.names = FALSE)

# Sample clustering (If specified in options)
hc_samples <- FALSE
if (cluster_samples) {
  if (distance_metric == "euclidean") {
    d_samples <- dist(t(mat_use), method = "euclidean")
  } else {
    cor_samples <- cor(mat_use, method = cor_method, use = "pairwise.complete.obs")
    d_samples <- as.dist(1 - cor_samples)
  }
  hc_samples <- hclust(d_samples, method = linkage_method)
}

# Heatmap + dendrogram
annotation_row <- data.frame(Cluster = factor(gene_clusters))
rownames(annotation_row) <- rownames(mat_use)

# Order clusters by the order in which they appear in the dendrogram to aid in visual interpretation
row_order <- hc_genes$order
ordered_genes <- rownames(mat_use)[row_order]
clusters_in_order <- gene_clusters[ordered_genes]

# Keep only first occurrence of each cluster
cluster_levels_ordered <- unique(clusters_in_order)

# Reorder the cluster
annotation_row$Cluster <- factor(
  gene_clusters,
  levels = cluster_levels_ordered
)

# Make gene cluster named by gene IDs
names(gene_clusters) <- rownames(mat_use)

# Set scale colors
clrsp <- colorRampPalette(c("blue", "white", "red"))
clrs <- clrsp(200)

# Custom scale
breaks <- seq(-2, 2, length.out = 201)

outfile <- paste0("./plots/AT2-samples_gene_hclust", "_heatmap_k", k, "_dist-", distance_metric, ".pdf")
pdf(outfile, width = 7, height = 9)
pheatmap(
  mat_use,
  cluster_rows = hc_genes,
  cluster_cols = hc_samples, # FALSE if cluster_samples == FALSE
  cutree_rows = k, # visually separates the k gene clusters
  annotation_row = annotation_row,
  # annotation_colors = annotation_colors,
  color = clrs,
  breaks = breaks,
  show_rownames = FALSE, # set TRUE if you have few genes and want names shown
  main = paste0(
    "Gene hclust (k=", k,
    ", linkage=", linkage_method,
    ", dist=", distance_metric,
    if (distance_metric == "correlation") paste0(" [", cor_method, "]") else "",
    if (scale_rows) ", row-zscore" else ""
  )
)
dev.off()
```